\documentclass[../thesis.tex]{subfiles}
\graphicspath{{\subfix{../figs/}}}
\begin{document}

\chapter{Characteristic words of the inferred cultural regions}
\label{ch:appendix_acr_words}

In \cref{fig:whole_data_clust}, we present the American cultural regions inferred from
our lexical analysis of Twitter data. We hereafter show the most characteristic words
for each region according to the specificity metric introduced in \cref{eq:spec}.
These words are classified into common lexical fields for the different cultural
regions. The lists are by no means exhaustive.

\begin{itemize}
  \item[\color{tol1}$\blacksquare$] Cluster 1
  \begin{itemize}
    \item[] \textbf{African-American variant forms and acronyms:} bih, frfr, stg, ian, otp, kno, ik, dnd, mfs, yo, sew, asf, wassup.
    \item[] \textbf{Fashion, music and food:} braids, waffle, grits, weave, lil, cookout, dreads, rappers, gucci.
    \item[] \textbf{Family and friends:} bruh, niggas, mama, dawg, shawty, auntie, folks, bae.
    \item[] \textbf{Location related:} atl, lsu, georgia, ga, carolina, saints, sc, falcons.
  \end{itemize}

  \item[\color{tol4}$\blacksquare$] Cluster 2
  \begin{itemize}
    \item[] \textbf{Sports:} win, teams, tourney, bracket, scoring, halftime, basketball, fouls, wrestling, innings, baseball.
    \item[] \textbf{Spectatorship:} cheering, announcers, congratulations, popcorn, hyped, concert, seats, autograph.
    \item[] \textbf{Teenage/school life:} girls, locker, choir, bffs, classroom, students, grader, alumni.
    \item[] \textbf{Location-related:} midwest, ohio, illinois, missouri, chicago, stl, minnesota, indiana, wisconsin, kc.
  \end{itemize}
  
  \item[\color{tol3}$\blacksquare$] Cluster 3
  \begin{itemize}
    \item[] \textbf{Natural features:} mountains, tree, snow, land, view, trail.
    \item[] \textbf{Outdoors activities:} adventures, spotter, trained, climb, trip.
    \item[] \textbf{Thoughts and emotions:} thought, think, cry, feel, crying, laughing, worry.
    \item[] \textbf{Personal communication:} said, understand, say, told, saying, heard, listen, trust, telling. (Importantly, these are negative features, i.e., it is a relatively low frequency of these words that is specific to the cluster.)
    \item[] \textbf{Location-related:} co, colorado, montana, utah, idaho, byu, mt, ut, va, wy.
  \end{itemize}
  
  \item[\color{tol2}$\blacksquare$] Cluster 4
  \begin{itemize}
    \item[] \textbf{Urban features:} freeway, homeless, traffic, ave.
    \item[] \textbf{Nationalities and origins:} puerto, dutch, latino, dominican, asian, immigrants, europe, rico, foreign, countries.
    \item[] \textbf{Violence:} violence, dangerous, attack, crime, insult, abusive, racist, threat.
    \item[] \textbf{Politics:} political, ban, nazi, indicted, protests, crisis, policies, presidency, elections, taxes, majority, obstruction.
    \item[] \textbf{Location related:} california, giants, nyc, ca, socal, cali, seattle, oregon, nj, sacramento, seahawks, sf, massachusetts, mets.
  \end{itemize}
  
  \item[\color{tol5}$\blacksquare$] Cluster 5
  \begin{itemize}
    \item[] \textbf{Mexican culture:} queso, mexican, tacos.
    \item[] \textbf{Variety features:} yalls, restroom, fixing, freaking, badass, highkey, fav, tripping.
    \item[] \textbf{Acronyms:} loml, sm, tbh, fml, idk, fm, fwm.
    \item[] \textbf{Emotions and relationships:} cry, heart, hurts, regret, friends, love, marry, boyfriend, laugh.
    \item[] \textbf{Location related:} whataburger, texas, tx, texans, dallas, cowboys, dfw, antonio, sa, spurs, mavs, as, ok, ou, okc, astros, rangers, austin.
  \end{itemize}
\end{itemize}

\end{document}
